---
title: "Data-Driven Segmentation of U.S. Colleges and Universities Using K-Means Clustering on IPEDS Data"
author: "Benjamin Minor"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
    link-citations: true
    link-bibliography: true
course: Capstone Projects in Data Science
bibliography: references.bib
nocite: "@*"
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} (Go to `slides.qmd`
to edit.)

## Introduction

Clustering is a central technique in unsupervised machine learning, used to
reveal natural groupings within complex datasets where class labels are
unknown. Among its methods, K-Means clustering remains widely applied for its
simplicity, scalability, and interpretability [@mohdtalib2023behavioral]. The
algorithm partitions data into \(k\) groups based on similarity, minimizing
within-cluster variance while increasing separation between clusters. This
structure is valuable when the goal is understanding rather than prediction,
such as comparing institutions by quantitative indicators of performance,
resources, and access.

In higher education research, cluster analysis has long been used to explore
structural and performance diversity across institutions [@huberty2005cluster].
This approach makes it possible to move beyond broad categorical labels and
examine measurable dimensions such as outcomes, funding, and enrollment scale.
A two-stage framework that combines factor analysis with clustering has also
been proposed to derive empirically grounded institutional types
[@elam2016classification].

Standardized data and documentation have increased the reliability of these
methods [@nces2023groupings]. Best practices for merging and cleaning
educational datasets emphasize reproducibility and schema consistency in
large-scale analysis [@kelchen2019merging]. Together, these sources point to
the value of structured, data-driven methods in institutional research. The
Integrated Postsecondary Education Data System (IPEDS), maintained by the
National Center for Education Statistics, provides the standardized foundation
for this type of large-scale institutional analysis. IPEDS collects consistent
and comparable data on U.S. colleges and universities, making it a central
source for empirical studies that analyze institutional differences through
quantitative modeling.

Hybrid models further advance peer identification by pairing quantitative
clustering with expert validation, improving interpretability and practical use
[@tair2022peer]. This balance between algorithmic structure and contextual
review supports robust benchmarking and decision-making.

The present study applies K-Means clustering to IPEDS data to identify
data-driven groupings of postsecondary institutions based on financial,
academic, and outcome indicators. The objective is to uncover underlying
institutional profiles that describe how U.S. colleges and universities align
or differ in measurable ways, offering a transparent and scalable framework for
understanding institutional diversity [@huberty2005cluster;
@nces2023groupings; @elam2016classification].

## Literature Review

<details>
<summary><strong>Source Summaries (click to expand)</strong></summary>

<div class="markdown" style="margin-top:1rem;">

**Proposal of a New Method to Classify Higher Education Institutions**  
[PDF Link – Otterbein University Digital Commons](https://digitalcommons.otterbein.edu/cgi/viewcontent.cgi?article=1031&context=stu_dist)  
Elam (2016) develops a two-step method that combines factor analysis and
cluster analysis to generate a new, data-driven classification of higher
education institutions. The study selects academic, financial, and demographic
variables, then uses clustering to uncover institutional types that better
reflect observed diversity than traditional labels. The results show how
quantitative segmentation can support more nuanced comparisons among colleges
and universities.

---

**Cluster Analysis of Multivariate Data: Efficiency Versus Interpretability of Classifications**  
[PDF Link – Biometrics (JSTOR/Publisher site)](https://doi.org/10.2307/2528559)  
Forgy (1965) is one of the early papers that lays out practical procedures for
cluster analysis in multivariate settings, including what became known as the
“Forgy initialization” for K-Means. The work discusses trade-offs between
statistical efficiency and interpretability when constructing groups from
continuous variables. It provides foundational guidance on how to select and
evaluate clustering structures, which informs later algorithmic choices in this
project.

---

**Cluster Analysis in Higher Education Research**  
[PDF Link – Springer](https://link.springer.com/content/pdf/10.1007/1-4020-3279-X_8.pdf)  
Huberty, Jordan, and Brandt (2005) review how cluster analysis has been used to
study structural diversity and performance patterns across higher education
institutions. They show that statistical grouping can reveal institutional
similarities that conventional categories obscure, especially when using
multiple financial and outcome measures. The chapter also emphasizes
methodological issues such as variable selection, scaling, and validation, all
of which directly shape the design of this capstone.

---

**Data Clustering: 50 Years Beyond K-Means**  
[PDF Link – Pattern Recognition Letters](https://doi.org/10.1016/j.patrec.2009.09.011)  
Jain (2010) surveys five decades of clustering research, using K-Means as a
starting point and comparing it to alternative algorithms and distance
measures. The paper highlights strengths of K-Means—speed, simplicity, and
interpretability—while also noting limitations around non-spherical clusters
and sensitivity to initialization. This broad context supports the decision to
use K-Means here while being clear about what kinds of structure it can and
cannot capture in institutional data.

---

**Merging Data to Facilitate Analyses**  
[PDF Link – Robert Kelchen](https://robertkelchen.com/wp-content/uploads/2019/11/kelchen_merging_original.pdf)  
Kelchen (2019) provides a practical guide to joining and cleaning education
datasets such as IPEDS and federal aid files. The report walks through common
problems—schema differences, missing values, duplicated identifiers, and
inconsistent labels—and recommends documentation practices for reproducible
research. This study follows those recommendations in how IPEDS components are
merged on `UNITID`, how derived variables (for example, revenue per FTE) are
constructed, and how missingness is handled.

---

**Least Squares Quantization in PCM**  
[PDF Link – IEEE Transactions on Information Theory](https://doi.org/10.1109/TIT.1982.1056489)  
Lloyd (1982) formalizes what is now widely known as the Lloyd algorithm for
K-Means clustering, framing it as a least-squares quantization problem. The
paper provides the iterative assignment-and-update procedure and shows that it
monotonically decreases the objective function, even though it may converge to
local minima. This work provides the mathematical basis for the K-Means
implementation used in the capstone and justifies the use of multiple random
starts.

---

**Identification of Student Behavioral Patterns in Higher Education Using K-Means Clustering and Support Vector Machine**  
[PDF Link – MDPI Applied Sciences](https://www.mdpi.com/2076-3417/13/5/3267)  
Mohd Talib et al. (2023) apply K-Means clustering to student behavior and
performance data, then use a Support Vector Machine to validate the resulting
groups. They show how normalized features and careful cluster validation can
reveal distinct learning behavior patterns that are actionable for instructors.
The study illustrates how unsupervised clustering can simplify complex
education datasets into interpretable segments, reinforcing the modeling
choices in this project.

---

**Institutional Groupings in IPEDS: Considerations for Data Use**  
[Web Link – NCES](https://nces.ed.gov/ipeds/use-the-data/institutional-groupings-in-ipeds)  
The NCES guidance on institutional groupings explains how IPEDS constructs and
documents sector, control, and other grouping variables. It stresses the need
for consistency and transparency when comparing institutions across official
groupings and derived peer sets. These recommendations motivate the use of
clear sector labels, harmonized finance fields, and reproducible code when
constructing the analytic file.

---

**A Hybrid Approach to Identifying National Peer Institutions**  
[PDF Link – Texas Association for Institutional Research (TAIR)](https://www.texas-air.org/wp-content/uploads/2022/03/E3-A-Hybrid-Approach-to-Identifying-National-Peer-Institutions.pdf)  
The TAIR report describes a hybrid approach that combines statistical clustering
with expert judgment to identify peer institutions. Initial peer groups are
generated from performance and enrollment data, then refined through local
review to improve face validity and usefulness. This work underscores that
cluster results gain value when paired with institutional context, a point that
frames this capstone’s clusters as a starting point for peer benchmarking
rather than a final classification.

</div>
</details>


## Methods

This study applies the **K-Means clustering algorithm** to group U.S. higher
education institutions into data-driven segments based on scale, resources,
access, and outcomes. The process follows the standard Lloyd–Forgy heuristic
for iterative centroid-based optimization, as described in the literature on
higher education clustering and institutional classification [@lloyd1982least].

---

### Objective and Notation

Let \(X \in \mathbb{R}^{n \times d}\) be the standardized feature matrix, where
each row represents an institution and each column a numeric feature (for
example, revenue per FTE, tuition, or completion rate). The goal of K-Means is
to partition \(n\) observations into \(k\) clusters
\(C = \{C_1, C_2, \ldots, C_k\}\) and determine centroids
\(\{\mu_1, \mu_2, \ldots, \mu_k\}\) that minimize the within-cluster sum of
squares (WCSS):

$$
\min_{C} \sum_{i=1}^{k} \sum_{x_j \in C_i} \|x_j - \mu_i\|^2
$$

For a fixed cluster assignment, the optimal centroid \(\mu_i\) is the
arithmetic mean of all observations in \(C_i\):

$$
\mu_i = \frac{1}{|C_i|} \sum_{x_j \in C_i} x_j
$$

This optimization problem is NP-hard in general, so the standard Lloyd–Forgy
iterative heuristic [@lloyd1982least] is used to approximate convergence
through alternating assignment and update steps.

---

### Lloyd’s Algorithm (Assignment and Update)

The algorithm alternates between two phases until it converges:

1. **Assignment step**  
   Each observation is assigned to the nearest cluster centroid based on
   Euclidean distance:
   $$
   C_i^{(t)} = \{x_j : \|x_j - \mu_i^{(t)}\|^2 \leq \|x_j - \mu_r^{(t)}\|^2,\ \forall r \in \{1, \ldots, k\}\}
   $$

2. **Update step**  
   After assignments, centroids are updated to reflect the mean of their
   current members:
   $$
   \mu_i^{(t+1)} = \frac{1}{|C_i^{(t)}|} \sum_{x_j \in C_i^{(t)}} x_j
   $$

The process repeats until there is minimal change in cluster assignments or
centroid locations between iterations [@lloyd1982least].

---

### Feature Scaling and Preprocessing

Because IPEDS variables differ in magnitude and measurement units (for example,
tuition in USD, completion rates as percentages, and revenue per FTE in
dollars), all continuous numeric variables were standardized using **z-score
normalization** prior to clustering [@lloyd1982least; @huberty2005cluster]:

$$
z = \frac{x - \bar{x}}{s}
$$

where \(\bar{x}\) is the sample mean and \(s\) is the sample standard deviation.
This ensures that each variable contributes on a comparable scale to the
Euclidean distances used in K-Means clustering [@lloyd1982least].

Categorical fields, such as institutional control type, were not standardized.
Instead, they were reserved for post-cluster interpretation and comparison.

To lessen the impact of extreme but valid institutional profiles (for example,
very large public systems or small for-profit colleges), **Winsorization** was
applied to cap values beyond the 1st and 99th percentiles. This preserves the
presence of these institutions while limiting their leverage on centroid
placement and cluster boundaries [@huberty2005cluster].

---

### Implementation Procedure

K-Means was implemented using R’s base `stats::kmeans()` function with multiple
random initializations (`nstart = 25`) to reduce sensitivity to initial
centroid placement. To ensure reproducibility, all random seeds were fixed
using `set.seed(2024)`.

Cluster inputs included:

- `REV_PER_FTE` (revenue per full-time equivalent)
- `INSTR_EXP_FTE` (instructional expenses per FTE)
- `TUITION1` (in-state undergraduate tuition)
- `OM8_PCT_COMP` (8-year completion rate)
- `RET_PCF` (retention rate, full-time; part-time as fallback)

The optimal number of clusters \(k\) was not pre-specified in this section. It
is evaluated empirically in the **Modeling and Results** section using the
Elbow method and Silhouette analysis, following
[@huberty2005cluster; @tair2022peer].

---

### Justification of Method

K-Means was selected for its interpretability and computational efficiency on
medium-dimensional data. Prior studies in higher education analytics have shown
that unsupervised clustering can reveal institutional typologies that
traditional classifications overlook [@huberty2005cluster;
@nces2023groupings; @mohdtalib2023behavioral; @elam2016classification]. This
approach supports empirical grouping based on shared performance and resource
characteristics, providing a basis for benchmarking and peer comparison
[@kelchen2019merging].

---

### Cluster Validation and Model Evaluation

To ensure that groupings derived from K-Means were both stable and
interpretable, multiple internal validation techniques were applied. Because
K-Means requires the number of clusters \(k\) to be specified in advance,
choosing an appropriate \(k\) is a key step that affects both interpretability
and robustness [@huberty2005cluster; @tair2022peer].

#### Elbow Method (Within-Cluster Sum of Squares)

The Elbow method evaluates the total within-cluster sum of squares (WCSS) across
a range of \(k\) values:

$$
WCSS(k) = \min_C \sum_{i=1}^{k} \sum_{x_j \in C_i} \|x_j - \mu_i\|^2
$$

As \(k\) increases, WCSS decreases because cluster radii shrink. The optimal
\(k\) is typically chosen at the “elbow” point: the inflection where additional
clusters yield diminishing reductions in WCSS. This heuristic balances compact
clusters with a parsimonious number of groups.

#### Silhouette Analysis

Silhouette analysis was used to complement the Elbow approach and assess how
well each institution fits within its assigned cluster compared to neighboring
clusters [@mohdtalib2023behavioral]. The Silhouette coefficient for observation
\(i\) is

$$
s(i) = \frac{b(i) - a(i)}{\max[a(i), b(i)]}
$$

where  

- \(a(i)\) is the average distance between \(i\) and all other points in its
  cluster (cohesion),  
- \(b(i)\) is the minimum average distance between \(i\) and points in the
  nearest neighboring cluster (separation).

Values range from \(-1\) to \(1\), with higher scores indicating better
separation and cohesion. The average Silhouette width was computed across all
institutions for each candidate \(k\), and the value of \(k\) with the highest
mean silhouette was treated as the most stable partition.

#### Model Stability and Reproducibility

To reduce randomness from centroid initialization, the clustering routine was
run with multiple random starts (`nstart = 25`) and a fixed seed
(`set.seed(2024)`), following best practices for reproducible analysis
[@kelchen2019merging]. This helps ensure that the resulting segmentation is
not an artifact of a single random state.

#### Summary

This validation framework combines geometric compactness (Elbow method) and
structural separation (Silhouette coefficient) to justify the final number of
clusters. The choice of \(k\) and the interpretation of the resulting segments
are presented in the following sections.

## Exploratory Data Analysis

### Data Exploration and Visualization

#### Dataset Overview

This study uses institutional-level data from the **Integrated Postsecondary
Education Data System (IPEDS)** for the 2023 collection year (2023–2024
cycle). IPEDS collects standardized information from all U.S. postsecondary
institutions that participate in federal financial aid programs, covering the
public, private nonprofit, and private for-profit sectors.

The following survey components were used and merged on the key `UNITID`:

- **HD2023** — Institutional characteristics (name, control, sector, etc.)
- **F2223_F1A** — Finance (public, GASB)
- **F2223_F2** — Finance (private nonprofit, FASB)
- **F2223_F3** — Finance (private for-profit)
- **EFIA2023** — 12-month instructional activity (source of FTE)
- **ADM2023** — Admissions (admission rate, tuition; used selectively)
- **OM2023** — Outcome measures (completion/attainment; used selectively)

All files were downloaded from the NCES IPEDS Data Center.

---

#### Variable Selection and Preparation

To represent institutional scale, financial capacity, access, and outcomes,
variables were drawn from multiple components and aligned by `UNITID`. The
main fields used are summarized below:

| Feature Group            | Variable (Field)                                  | Description |
|--------------------------|---------------------------------------------------|-------------|
| **Institutional Scale**  | `FTE_TOTAL` (built from EFIA2023)                | Total full-time-equivalent enrollment (UG + Grad + DPP if present). |
| **Financial Resources**  | `TOT_REV` (F1D01 / F2D162 / F3D12)               | Total revenues (public), total revenues and investment return (nonprofit/for-profit). |
|                          | `INSTR_EXP` (F1C011 / F2E011 / F3E011)           | Instructional expenses (aligned across control types). |
|                          | `TOT_EXP` (F1N07 / F2B02 / F3B02)                | Total operating expenses. |
| **Access and Price**     | `ADM_RATE`, `TUITION1` (ADM2023 / IC2023)        | Admission rate; in-state tuition and fees. |
| **Outcomes**             | `OM8_PCT_COMP` (OM2023 or derived)               | 8-year outcome/completion measure. |

Where variables differed by control type, fields were harmonized to shared
names (for example, `TOT_REV`, `INSTR_EXP`, `TOT_EXP`). Ratios per FTE were
computed after merging FTE data.

---

#### Data Cleaning and Integration

Finance data from public (`F1A`), private nonprofit (`F2`), and private
for-profit (`F3`) files were imported and mapped to total revenue,
instructional expense, and total expense. Twelve-month instructional activity
(`EFIA2023`) provided undergraduate, graduate, and total FTE values; reported
FTE were preferred, with estimated values as fallback. Monetary values are
reported in nominal USD for 2023–24 and are not inflation-adjusted.

Tables were merged with left joins on `UNITID` to preserve coverage. Numeric
missing values were left as `NA` in order to stay consistent with NCES
reporting. After filtering to active, degree-granting institutions with valid
FTE, the analytic file contains roughly 6,000 institutions across all three
control types.

```{r}
# ----libraries-------------------------------------------------------------
library(tidyverse)   # data wrangling / joins / ggplot2
library(janitor)     # clean column names
library(gt)          # summary tables
library(scales)      # formatting helpers
library(GGally)      # quick pair plots (optional)
library(corrr)       # correlation summaries (optional)
library(here)        # reproducible relative paths
library(patchwork)   # plot layouts (optional)

# ----paths-----------------------------------------------------------------
data_dir <- here("data")  # change if your files live elsewhere

# ----helper: robust numeric coercion--------------------------------------
numify <- function(x) as.numeric(gsub("[^0-9.\\-]", "", as.character(x)))

# ----load core IPEDS components-------------------------------------------
hd  <- read_csv(file.path(data_dir, "HD2023.csv"))   |> clean_names()
adm <- read_csv(file.path(data_dir, "ADM2023.csv"))  |> clean_names()
om  <- read_csv(file.path(data_dir, "OM2023.csv"))   |> clean_names()

# Finance (public / private nonprofit / private for-profit)
f1a <- read_csv(file.path(data_dir, "F2223_F1A.csv")) |> clean_names()
f2  <- read_csv(file.path(data_dir, "F2223_F2.csv"))  |> clean_names()
f3  <- read_csv(file.path(data_dir, "F2223_F3.csv"))  |> clean_names()

# 12-month instructional activity (FTE source)
efia <- read_csv(
  file.path(data_dir, "EFIA2023.csv"),
  na = c("", ".", "-", "—", "NA", "N/A")
) |> clean_names()

# Student charges (tuition and fees)
ic_ay <- read_csv(file.path(data_dir, "IC2023.csv"))  |> clean_names()

# Student financial aid (Pell Grant %)
sfa <- read_csv(file.path(data_dir, "SFA2223.csv"))   |> clean_names()

# Fall retention & student-faculty ratio
ef2023d <- read_csv(file.path(data_dir, "EF2023D.csv")) |> clean_names()

# ----EFIA -> FTE totals (coerce safely)-----------------------------------
efia <- efia |>
  mutate(across(any_of(c("fteug","efteug","ftegd","eftegd","ftedpp")), numify))

efia_fte <- efia |>
  transmute(
    unitid,
    fte_ug    = coalesce(fteug,  efteug),
    fte_grad  = coalesce(ftegd,  eftegd),
    fte_dpp   = ftedpp,
    fte_total = coalesce(fteug,  efteug, 0) +
                coalesce(ftegd,  eftegd, 0) +
                coalesce(ftedpp, 0)
  )

# ----finance: align to common names & coerce numeric-----------------------
f1a_fin <- f1a |>
  mutate(across(any_of(c("f1d01","f1c011","f1n07")), numify)) |>
  transmute(
    unitid,
    tot_rev   = f1d01,
    instr_exp = f1c011,
    tot_exp   = f1n07,
    control   = "Public"
  )

f2_fin <- f2 |>
  mutate(across(any_of(c("f2d162","f2e011","f2b02")), numify)) |>
  transmute(
    unitid,
    tot_rev   = f2d162,
    instr_exp = f2e011,
    tot_exp   = f2b02,
    control   = "Private Nonprofit"
  )

f3_fin <- f3 |>
  mutate(across(any_of(c("f3d12","f3e011","f3b02")), numify)) |>
  transmute(
    unitid,
    tot_rev   = f3d12,
    instr_exp = f3e011,
    tot_exp   = f3b02,
    control   = "Private For-Profit"
  )

fin <- bind_rows(f1a_fin, f2_fin, f3_fin)

# ----unified sector handling-----------------------------------------------
fin <- fin |>
  mutate(sector_fin = control) |>
  select(-control)

if ("control" %in% names(hd)) {
  hd <- hd |>
    mutate(
      sector_hd = dplyr::case_match(
        control,
        1 ~ "Public",
        2 ~ "Private Nonprofit",
        3 ~ "Private For-Profit",
        .default = NA_character_
      )
    )
} else {
  hd <- hd |> mutate(sector_hd = NA_character_)
}

# ----tuition, Pell, retention, stufacr (robust numeric)-------------------
ic_sel <- ic_ay |>
  mutate(across(any_of(c("tuition1")), numify)) |>
  transmute(unitid, tuition1)

sfa_sel <- sfa |>
  mutate(across(any_of(c("pgrnt_p")), numify)) |>
  transmute(unitid, pct_pell = pgrnt_p)

ef_ret <- ef2023d |>
  mutate(across(any_of(c("ret_pcf","ret_pcp","stufacr")), numify)) |>
  transmute(unitid, ret_pcf, ret_pcp, stufacr)

# ----merge all components (single clean 'sector')--------------------------
ipeds <- hd |>
  left_join(fin,      by = "unitid") |>
  left_join(efia_fte, by = "unitid") |>
  left_join(ic_sel,   by = "unitid") |>
  left_join(sfa_sel,  by = "unitid") |>
  left_join(ef_ret,   by = "unitid") |>
  left_join(adm,      by = "unitid") |>
  left_join(om,       by = "unitid") |>
  mutate(
    sector = dplyr::coalesce(sector_hd, sector_fin),
    sector = factor(
      sector,
      levels = c("Public", "Private Nonprofit", "Private For-Profit")
    )
  ) |>
  filter(!is.na(unitid), !is.na(fte_total), fte_total > 0) |>
  distinct(unitid, .keep_all = TRUE) |>
  mutate(
    rev_per_fte    = tot_rev   / fte_total,
    instr_exp_fte  = instr_exp / fte_total,
    exp_per_fte    = tot_exp   / fte_total,
    retention_rate = coalesce(ret_pcf, ret_pcp)
  )

# ----quick composition summary---------------------------------------------
ipeds |>
  summarise(
    total_institutions = n(),
    public_share       = mean(sector == "Public",             na.rm = TRUE),
    nonprofit_share    = mean(sector == "Private Nonprofit",  na.rm = TRUE),
    forprofit_share    = mean(sector == "Private For-Profit", na.rm = TRUE)
  ) |>
  gt() |>
  fmt_percent(columns = 2:4, decimals = 1)

# ----prep for visuals / modeling------------------------------------------
if (!"om8_pct_comp" %in% names(ipeds)) {
  need <- c("omawdp8","omnoawd","omenrap","omenryp")
  if (all(need %in% names(ipeds))) {
    ipeds <- ipeds |>
      mutate(
        om_total     = omawdp8 + omnoawd + omenrap + omenryp,
        om8_pct_comp = if_else(om_total > 0, 100 * omawdp8 / om_total, NA_real_)
      )
  }
}

if ("tuition1" %in% names(ipeds) && !is.numeric(ipeds$tuition1)) {
  ipeds <- ipeds |>
    mutate(tuition1 = readr::parse_number(as.character(tuition1)))
}

vis_df <- ipeds |>
  filter(!is.na(sector),
         !is.na(tuition1),
         !is.na(rev_per_fte),
         !is.na(instr_exp_fte),
         !is.na(om8_pct_comp)) |>
  mutate(
    tuition_k = tuition1 / 1000,
    rev_fte_k = rev_per_fte / 1000,
    exp_fte_k = instr_exp_fte / 1000
  )
```

#### Figure 1. Distributions of Core Metrics

```{r}
# ----figure-1--------------------------------------------------------------
#| label: fig-dist
#| fig-cap: "Distributions of Key Variables"
vis_df |>
  mutate(
    tuition_k = tuition1 / 1000,
    rev_fte_k = rev_per_fte / 1000,
    exp_fte_k = instr_exp_fte / 1000
  ) |>
  select(tuition_k, rev_fte_k, exp_fte_k, om8_pct_comp) |>
  pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "#4682B4", color = "white", alpha = 0.85) +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  labs(
    title = "Figure 1. Distributions of Key Variables",
    subtitle = "Spread and skew across tuition, finance per FTE, and completion rates",
    x = NULL, y = "Count"
  )
```

#### Figure 2. Tuition vs Completion by Sector

```{r}
#----figure-2--------------------------------------------------------------
#| label: fig-tuition-completion
#| fig-cap: "Tuition vs 8-Year Completion by Sector"
ggplot(vis_df, aes(tuition1, om8_pct_comp, color = sector)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "gray40") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Figure 2. Tuition vs 8-Year Completion by Sector",
    subtitle = "Dashed global fit shows a modest positive relationship",
    x = "In-state tuition and fees (USD)",
    y = "Completion rate (8-year, %)",
    color = "Sector"
  )
```

#### Figure 3. Revenue vs Instructional Expense per FTE

```{r}
#----figure-3--------------------------------------------------------------
#| label: fig-rev-exp-fte
#| fig-cap: "Financial Profile per FTE"
vis_df |>
  mutate(
    rev_fte_k = rev_per_fte / 1000,
    exp_fte_k = instr_exp_fte / 1000
  ) |>
  ggplot(aes(rev_fte_k, exp_fte_k, color = sector)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Figure 3. Financial Profile per FTE",
    subtitle = "Public institutions cluster tightly; nonprofits span wider ranges; for-profits leaner on expense",
    x = "Revenue per FTE (thousands USD)",
    y = "Instructional expense per FTE (thousands USD)",
    color = "Sector"
  )
```

#### Figure 4. Correlation of Key Variables

```{r}
#----figure-4--------------------------------------------------------------
#| label: fig-corr
#| fig-cap: "Correlation Heatmap of Tuition, Finances, and Outcomes"
vis_df |>
  select(tuition1, rev_per_fte, instr_exp_fte, om8_pct_comp) |>
  correlate(method = "pearson", use = "pairwise.complete.obs") |>
  stretch(na.rm = TRUE) |>
  ggplot(aes(x, y, fill = r)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    limits = c(-1, 1), name = "r"
  ) +
  labs(
    title = "Figure 4. Correlation Heatmap of Tuition, Finances, and Outcomes",
    subtitle = "Finance measures move together and relate positively to outcomes",
    x = NULL, y = NULL
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

#### Figure 5. Student-to-Faculty Ratio by Sector

```{r}
#----figure-5--------------------------------------------------------------
pal_sector <- c(
  "Public"             = "#E15759",
  "Private Nonprofit"  = "#59A14F",
  "Private For-Profit" = "#4E79A7"
)

ipeds |>
  filter(!is.na(sector), !is.na(stufacr), stufacr > 0) |>
  mutate(stufacr_cap = pmin(stufacr, quantile(stufacr, 0.99, na.rm = TRUE))) |>
  ggplot(aes(y = sector, x = stufacr_cap, fill = sector)) +
  geom_boxplot(width = 0.5, outlier.alpha = 0.25, outlier.size = 0.9) +
  scale_fill_manual(values = pal_sector, guide = "none") +
  labs(
    title    = "Figure 5. Student-to-Faculty Ratio by Sector",
    subtitle = "Public institutions show tighter ratios; private institutions vary more widely",
    x        = "Student-to-faculty ratio",
    y        = NULL
  ) +
  theme(legend.position = "none")
```

#### Figure 6. Retention vs Completion by Sector

```{r}
#----figure-6--------------------------------------------------------------
ipeds |>
  filter(!is.na(sector), !is.na(retention_rate), !is.na(om8_pct_comp)) |>
  ggplot(aes(x = retention_rate, y = om8_pct_comp, color = sector)) +
  geom_point(alpha = 0.55, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "gray40") +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Figure 6. Retention vs Completion by Sector",
    subtitle = "Higher retention rates align strongly with higher completion rates across all sectors",
    x = "First-year retention rate (%, full-time preferred; else part-time)",
    y = "Completion rate (8-year, %)",
    color = "Sector"
  )
```

## Modeling and Results

This section presents the analytical process and results of applying the
K-Means clustering algorithm to institutional data from IPEDS. While the
*Methods* section laid out the modeling framework and variable design, this
section focuses on how the model was implemented, how the optimal number of
clusters was chosen, and what structural patterns emerged across institutional
segments.

---

### Data Preparation for Modeling

Prior to clustering, all selected quantitative features were standardized using
z-score normalization to ensure comparability across scales. The final modeling
dataset included institutions with valid and complete entries for key
financial, access, and outcome metrics:

- `REV_PER_FTE`: revenue per full-time equivalent student  
- `INSTR_EXP_FTE`: instructional expense per FTE  
- `TUITION1`: in-state undergraduate tuition and fees  
- `OM8_PCT_COMP`: 8-year student completion rate  
- `RETENTION_RATE`: first-year retention (full-time preferred; part-time if missing)

Each row of the modeling dataset corresponds to a unique institution (`UNITID`),
and each column represents a standardized feature describing resources,
affordability, or student outcomes. To reduce the influence of extreme but
valid outliers (for example, very large public systems or small proprietary
colleges), variables were Winsorized at the 1st and 99th percentiles. This
preserves these institutions in the data while limiting their impact on
centroid placement.

---

### Determining the Optimal Number of Clusters

Selecting an appropriate number of clusters, \(k\), is a key decision in
K-Means because it determines the level of segmentation and affects both
interpretability and stability. To guide this choice, two internal validation
techniques were applied:

1. **Elbow method (within-cluster sum of squares, WCSS)** — used to identify
   the point where adding more clusters yields diminishing improvements in
   compactness.  
2. **Silhouette analysis** — used to evaluate how well each institution fits
   within its assigned cluster relative to neighboring clusters.

Together, these diagnostics balance parsimony (fewer, more interpretable
clusters) and stability (coherent and well-separated clusters). Both methods
pointed toward \(k \approx 4\) as a stable and interpretable solution.

```{r}
#| label: elbow-silhouette
#| fig-cap: "Elbow and Silhouette Analysis for Optimal k"
#| echo: true
#| message: false
#| warning: false

library(cluster)

# Prepare modeling matrix
X <- vis_df |>
  select(rev_per_fte, instr_exp_fte, tuition1, om8_pct_comp, retention_rate) |>
  drop_na() |>
  scale()

# --- Elbow ---------------------------------------------------------
set.seed(2024)
wcss <- map_dbl(1:10, ~ kmeans(X, centers = .x, nstart = 25)$tot.withinss)
elbow_df <- tibble(k = 1:10, wcss = wcss)

ggplot(elbow_df, aes(k, wcss)) +
  geom_line(color = "#4682B4", linewidth = 1) +
  geom_point(size = 2, color = "#B2182B") +
  geom_vline(xintercept = 4, linetype = "dashed", color = "gray50") +
  annotate("text", x = 4.25, y = max(wcss)*0.8, label = "Elbow ≈ 4", size = 3.2, hjust = 0) +
  labs(
    title = "Figure 7. Elbow Method for Optimal k",
    subtitle = "Elbow point suggests diminishing returns beyond k = 4",
    x = "Number of clusters (k)",
    y = "Total within-cluster sum of squares (WCSS)"
  ) +
  theme_minimal(base_size = 11)

# --- Silhouette -----------------------------------------------------
silhouette_scores <- map_dbl(2:10, function(k) {
  km <- kmeans(X, centers = k, nstart = 25)
  ss <- silhouette(km$cluster, dist(X))
  mean(ss[, 3])
})

sil_df <- tibble(k = 2:10, silhouette = silhouette_scores)

ggplot(sil_df, aes(k, silhouette)) +
  geom_line(color = "#4B9CD3", linewidth = 1) +
  geom_point(size = 2, color = "#E66101") +
  geom_vline(xintercept = 4, linetype = "dashed", color = "gray50") +
  annotate("text", x = 4.3, y = max(silhouette_scores)*0.97, label = "Peak ≈ 4", size = 3.2, hjust = 0) +
  labs(
    title = "Figure 8. Silhouette Analysis for Cluster Cohesion and Separation",
    subtitle = "Peak silhouette width supports k = 4",
    x = "Number of clusters (k)",
    y = "Average silhouette width"
  ) +
  theme_minimal(base_size = 11)
```

### Final Clustering Model (k = 4)

```{r}
#| label: final-kmeans
#| fig-cap: "K-Means Segmentation Results (k = 4)"
#| echo: true
#| message: false
#| warning: false

set.seed(2024)
final_k <- 4

model_df <- vis_df |>
  select(rev_per_fte, instr_exp_fte, tuition1, om8_pct_comp, retention_rate) |>
  drop_na()

X <- scale(model_df)
k_model <- kmeans(X, centers = final_k, nstart = 25)

ipeds_clustered <- model_df |>
  mutate(cluster = factor(k_model$cluster))

cluster_summary <- ipeds_clustered |>
  group_by(cluster) |>
  summarise(
    avg_tuition    = mean(tuition1,       na.rm = TRUE),
    avg_rev_fte    = mean(rev_per_fte,    na.rm = TRUE),
    avg_exp_fte    = mean(instr_exp_fte,  na.rm = TRUE),
    avg_completion = mean(om8_pct_comp,   na.rm = TRUE),
    avg_retention  = mean(retention_rate, na.rm = TRUE),
    n              = n()
  )

cluster_summary |>
  gt() |>
  fmt_currency(
    columns = c(avg_tuition, avg_rev_fte, avg_exp_fte),
    decimals = 0
  ) |>
  fmt_percent(
    columns = c(avg_completion, avg_retention),
    scale_values = FALSE,
    decimals = 1
  ) |>
  tab_header(
    title = "Table 1. Cluster-Level Summary Statistics",
    subtitle = "Average tuition, resources, and outcomes across institutional clusters"
  )
```

```{r}
#| label: fig-cluster-counts
#| fig-cap: "Distribution of Institutions Across Clusters (k = 4)"
#| echo: true

ipeds_clustered |>
  ggplot(aes(x = cluster, fill = cluster)) +
  geom_bar(alpha = 0.9) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "Cluster",
    y = "Count of institutions",
    title = "Figure 9. Distribution of Institutions Across Clusters",
    subtitle = "Overall size of each institutional group under k = 4"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

#### Diagnostic Check: Variable Distributions by Cluster

```{r}
#| label: boxplots-cluster-vars
#| fig-cap: "Distribution of Standardized Variables Across Clusters"
#| echo: false

ipeds_clustered |>
  pivot_longer(
    cols = c(rev_per_fte, instr_exp_fte, tuition1, om8_pct_comp, retention_rate),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = cluster, y = value, fill = cluster)) +
  geom_boxplot(alpha = 0.85, outlier.alpha = 0.3) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Figure 10. Distribution of Variables Across Clusters",
    x = "Cluster",
    y = "Value (standardized)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

### Cluster Segmentation Narratives

#### Cluster 1 — Broad-Access, Lower-Resource Institutions

This segment tends to include institutions with lower tuition, limited
instructional spending, and below-average completion rates. They are often
regional, commuter-serving, or open-access colleges where affordability is
high, but resource constraints can affect instructional capacity and student
support. Retention is modest, indicating challenges in sustaining student
persistence.

#### Cluster 2 — Moderately Resourced Institutions with Balanced Outcomes

Cluster 2 exhibits mid-range tuition, moderate spending, and solid student
outcomes. These institutions represent a balanced “middle tier” of higher
education, often including regional publics and private nonprofits with stable
performance metrics despite lower resource levels than elite institutions.

#### Cluster 3 — High-Tuition, High-Resource Institutions

Institutions in this segment show substantially higher tuition, significant
instructional investment, and correspondingly high completion and retention
rates. This profile aligns with selective universities and well-resourced
private institutions that operate on a “high input, high output” model.

#### Cluster 4 — Low-Tuition but Strong-Outcome Institutions

This segment includes institutions that maintain lower-than-average tuition yet
achieve strong completion and retention rates. Often efficient public
institutions or community colleges, these schools show that strong student
outcomes do not always require high expenditure levels.

### Visualizing Cluster Structure in Feature Space

To visualize how institutions separate in high-dimensional space, the
standardized modeling matrix was projected into two components using Principal
Component Analysis (PCA). PCA was not used for model fitting; instead, it
provides a compact visualization that shows whether clusters form coherent
regions. PC1 is primarily associated with overall resource intensity, and PC2
captures variation related to affordability and outcomes.

```{r}
#| label: fig-pca-clusters
#| fig-cap: "PCA Visualization of K-Means Clusters"
#| echo: true

pca_res <- prcomp(X, scale. = FALSE)

pca_df <- as_tibble(pca_res$x[, 1:2]) |>
  mutate(cluster = ipeds_clustered$cluster)

ggplot(pca_df, aes(PC1, PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Figure 11. PCA Visualization of K-Means Clusters",
    subtitle = "Clusters separate clearly across the first two principal components",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal(base_size = 11)
```

## Conclusion

This project used K-Means clustering on IPEDS data to break institutions into
groups based on resources, affordability, and student success. The variables
were standardized and Winsorized to keep outliers from pulling the model
around. Both the Elbow method and the Silhouette scores pointed toward four
clusters, and that setup ended up giving the clearest and most stable
structure.

The final groups showed four different types of institutions. One group
captured broad access schools that operate with fewer resources and see lower
retention and completion. Another group sat in the middle, with moderate
spending, moderate tuition, and steady outcomes. A third group represented
well-funded schools with high tuition and strong performance. The last group
stood out because they kept tuition low while still producing solid retention
and completion. That pattern shows that resources matter, but they are not the
whole story when it comes to student outcomes.

The diagnostic checks backed this up. The boxplots showed clear separation
across each variable, and the PCA plot showed four distinct groups when the
data was compressed into two components. PCA was only used to visualize
structure, but it made the spacing between clusters easier to see.

Overall, the results show that a small, focused set of metrics is enough to
reveal real structure in the higher education landscape. These segments can
help with peer grouping, benchmarking, and conversations about funding and
equity. Future work could bring in more variables, test other clustering
approaches, or look at how student demographics and program mix change the
picture. Even with the limited scope here, the model gives a clear and useful
view of how institutions differ across the system.

## References
