---
title: "Data-Driven Segmentation of U.S. Colleges and Universities Using K-Means Clustering on IPEDS Data"
author: "Benjamin Minor"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
    link-citations: true       # <-- move here
    link-bibliography: true    # <-- and here
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
nocite: "@*"
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)



## Introduction

Clustering is a central technique in unsupervised machine learning, designed to reveal natural groupings within complex datasets where class labels are unknown. Among its methods, K-Means clustering remains widely applied for its simplicity, scalability, and interpretability [@mohdtalib2023behavioral]. The algorithm partitions data into *k* groups based on similarity, minimizing within-cluster variance while maximizing between-cluster separation. This structure is valuable when the goal is understanding rather than prediction, such as comparing institutions by quantitative indicators of performance, resources, and access.

In higher education research, cluster analysis has long been used to explore structural and performance diversity across institutions [@huberty2005cluster]. This approach enables analyses that move beyond categorical labels and examine measurable dimensions such as outcomes, funding, and enrollment scale. A two-stage framework that combines factor analysis with clustering has also been proposed to derive empirically grounded institutional types [@elam2016classification].

Standardized data and documentation have increased the reliability of these methods [@nces2023groupings]. Best practices for merging and cleaning educational datasets emphasize reproducibility and schema consistency in large-scale analysis [@kelchen2019merging]. Together, these sources point to the value of structured, data-driven methods in institutional research. The Integrated Postsecondary Education Data System (IPEDS), maintained by the National Center for Education Statistics, provides the standardized foundation for this type of large-scale institutional analysis. IPEDS collects consistent and comparable data on U.S. colleges and universities, making it a central source for empirical studies that analyze institutional differences through quantitative modeling.

Hybrid models further advance peer identification by pairing quantitative clustering with expert validation, improving interpretability and practical use [@tair2022peer]. This balance between algorithmic precision and contextual review supports robust benchmarking and decision-making.

The present study applies K-Means clustering to IPEDS data to identify data-driven groupings of postsecondary institutions based on financial, academic, and outcome indicators. The objective is to uncover underlying institutional profiles that describe how U.S. colleges and universities align or differ in measurable ways, offering a transparent and scalable framework for understanding institutional diversity [@huberty2005cluster; @nces2023groupings; @elam2016classification].


## Literature Review

<details>
<summary><strong>Source Summaries (click to expand)</strong></summary>

<div class="markdown" style="margin-top:1rem;">

**Cluster Analysis in Higher Education Research**  
[PDF Link – Springer](https://link.springer.com/content/pdf/10.1007/1-4020-3279-X_8.pdf)  
The chapter reviews how cluster analysis is used to study diversity and structure among higher education institutions. The authors explain how statistical grouping exposes institutional similarities that typical classification systems overlook. They discuss the method’s flexibility for understanding performance and policy patterns across universities and show how it can support empirical research on institutional behavior.  

---

**Institutional Groupings in IPEDS: Considerations for Data Use**  
[Web Link – NCES](https://nces.ed.gov/ipeds/use-the-data/institutional-groupings-in-ipeds)  
This NCES publication outlines the logic and procedures for creating institutional groupings using standardized IPEDS variables. It focuses on the need for consistency, variable alignment, and transparency when comparing data across schools. The document highlights how structured variable relationships can support reproducible and fair institutional comparisons across the national education system.  

---

**Identification of Student Behavioral Patterns in Higher Education Using K-Means Clustering and Support Vector Machine**  
[PDF Link – MDPI Applied Sciences](https://www.mdpi.com/2076-3417/13/5/3267)  
The study applies K-Means clustering to student behavior and performance data, validating results with a Support Vector Machine. The authors used normalized features, centroid checks, and cluster validation to identify three distinct learning behavior groups. They found that clustering helped simplify large datasets into actionable patterns, showing how unsupervised models can classify educational behavior at scale.  

---

**Proposal of a New Method to Classify Higher Education Institutions**  
[PDF Link – Otterbein University Digital Commons](https://digitalcommons.otterbein.edu/cgi/viewcontent.cgi?article=1031&context=stu_dist)  
This thesis introduces a two-step method that combines factor analysis and cluster analysis to build a new classification of universities. The author selected academic, financial, and demographic indicators to find groupings that better represent institutional diversity. The results show that data-driven group formation can outperform fixed classification systems and highlight how numerical traits can define institutional identity.  

---

**A Hybrid Approach to Identifying National Peer Institutions**  
[PDF Link – Texas Association for Institutional Research (TAIR)](https://www.texas-air.org/wp-content/uploads/2022/03/E3-A-Hybrid-Approach-to-Identifying-National-Peer-Institutions.pdf)  
This report explains how statistical clustering and expert review can be combined to identify peer institutions. Quantitative results were first generated from performance and enrollment data, then refined through institutional judgment to improve interpretability. The authors conclude that pairing data models with professional context leads to more balanced and credible peer comparisons.  

---

**Merging Data to Facilitate Analyses**  
[PDF Link – Robert Kelchen](https://robertkelchen.com/wp-content/uploads/2019/11/kelchen_merging_original.pdf)  
Kelchen describes the process of joining and cleaning educational data for large-scale analysis. The guide details how to handle schema differences, missing values, and inconsistent labels across datasets like IPEDS and federal records. It stresses the importance of reproducibility and documentation when merging data sources to maintain accuracy in quantitative research.  


</div>
</details>

## Methods

## Analysis and Results

### Data Description and Visualization

#### Dataset Overview

This study uses institutional-level data from the **Integrated Postsecondary
Education Data System (IPEDS)** for the 2023 collection year (2023–2024
cycle). IPEDS collects standardized information from all U.S. postsecondary
institutions participating in federal financial aid programs, covering public,
private nonprofit, and private for-profit sectors.

The following survey components were used and merged on the key **`UNITID`**:

- **HD2023** — Institutional characteristics (name, control, sector, etc.)
- **F2223_F1A** — Finance (public, GASB)
- **F2223_F2** — Finance (private nonprofit, FASB)
- **F2223_F3** — Finance (private for-profit)
- **EFIA2023** — 12-month instructional activity (**source of FTE**)
- **ADM2023** — Admissions (admission rate, tuition; optional in analysis)
- **OM2023** — Outcome measures (completion/attainment; optional in analysis)

All files were downloaded from the NCES IPEDS Data Center.

---

#### Variable Selection and Preparation

To represent institutional scale, financial capacity, access, and
outcomes, variables were drawn from multiple components and aligned by
`UNITID`. Below are the primary fields used in this phase:

| Feature Group          | Variable (Field)                                  | Description |
|---|---|---|
| **Institutional Scale** | `FTE_TOTAL` (built from EFIA2023)                 | Total full-time-equivalent enrollment (UG + Grad + DPP if present). |
| **Financial Resources** | `TOT_REV` (F1D01 / F2D162 / F3D12)                 | Total revenues (public), total revenues & investment return (nonprofit/for-profit). |
|                          | `INSTR_EXP` (F1C011 / F2E011 / F3E011)            | Instructional expenses (aligned across control types). |
|                          | `TOT_EXP` (F1N07 / F2B02 / F3B02)                 | Total operating expenses. |
| **Access & Price**       | e.g., `ADM_RATE`, `TUITION1` (ADM2023)            | Admission rate; in-state tuition/fees (if used). |
| **Outcomes**             | e.g., `OM2_PCT_COMP` (OM2023)                     | 8-year outcome/completion measure (if used). |

> Note: Where variables differ by control type, fields were harmonized to a
> shared name (e.g., `TOT_REV`, `INSTR_EXP`, `TOT_EXP`). Ratios per FTE are
> derived after FTE merge.

---

#### Data Cleaning and Integration

Finance data from public (`F1A`), private nonprofit (`F2`), and private
for-profit (`F3`) files were imported and standardized to total revenue,
instructional expense, and total expense. Twelve-month instructional
activity (`EFIA2023`) provided undergraduate, graduate, and total FTE
values; reported FTE were preferred, with estimated values as fallback.

Tables were merged using left joins on `UNITID` to preserve coverage. Numeric
missing values were left as `NA` to maintain fidelity with NCES reporting.
After filtering to active, degree-granting institutions with valid FTE, the
analytic file contains ~6,000 institutions across all three control types.

```{r}
# ----libraries-------------------------------------------------------------
library(tidyverse)   # Data wrangling/joins/plots
library(janitor)     # Clean column names
library(gt)          # Summary tables
library(scales)      # Formatting
library(GGally)      # Quick pair plots
library(corrr)       # Correlation summaries
library(here)        # Reproducible relative paths

# ----directory and base imports-------------------------------------------
data_dir <- here("data")  # adjust if needed

# Core IPEDS components
hd  <- read_csv(file.path(data_dir, "HD2023.csv"))  |> clean_names()
adm <- read_csv(file.path(data_dir, "ADM2023.csv")) |> clean_names()
om  <- read_csv(file.path(data_dir, "OM2023.csv"))  |> clean_names()

# Finance (public / private nonprofit / private for-profit)
f1a <- read_csv(file.path(data_dir, "F2223_F1A.csv")) |> clean_names()
f2  <- read_csv(file.path(data_dir, "F2223_F2.csv"))  |> clean_names()
f3  <- read_csv(file.path(data_dir, "F2223_F3.csv"))  |> clean_names()

# 12-month instructional activity (FTE source)
efia <- read_csv(file.path(data_dir, "EFIA2023.csv")) |> clean_names()

# Student charges (tuition and fees)
ic_ay <- read_csv(file.path(data_dir, "IC2023.csv")) |> clean_names()

# Student financial aid (Pell Grant recipients)
sfa <- read_csv(file.path(data_dir, "SFA2223.csv")) |> clean_names()

# ----build FTE from EFIA2023----------------------------------------------
efia_fte <- efia |>
  mutate(across(c(fteug, efteug, ftegd, eftegd, ftedpp),
                ~ readr::parse_number(as.character(.)))) |>
  transmute(
    unitid,
    fte_ug    = coalesce(fteug, efteug),
    fte_grad  = coalesce(ftegd, eftegd),
    fte_dpp   = ftedpp,
    fte_total = coalesce(fteug, efteug, 0) +
                coalesce(ftegd, eftegd, 0) +
                coalesce(ftedpp, 0)
  )

# ----align finance variables across control types-------------------------
f1a_fin <- f1a |>
  transmute(
    unitid,
    tot_rev   = f1d01,   # Total revenues & other additions (public)
    instr_exp = f1c011,  # Instruction - current year total
    tot_exp   = f1n07,   # Total expenses
    control   = "Public"
  )

f2_fin <- f2 |>
  transmute(
    unitid,
    tot_rev   = f2d162,  # Total revenues & investment return - unrestricted
    instr_exp = f2e011,  # Instruction - total amount
    tot_exp   = f2b02,   # Total expenses
    control   = "Private Nonprofit"
  )

f3_fin <- f3 |>
  transmute(
    unitid,
    tot_rev   = f3d12,   # Total revenues & investment return
    instr_exp = f3e011,  # Instruction - total amount
    tot_exp   = f3b02,   # Total expenses
    control   = "Private For-Profit"
  )

fin <- bind_rows(f1a_fin, f2_fin, f3_fin)

# ----select tuition and Pell variables------------------------------------

ic_sel <- ic_ay |>
  transmute(
    unitid,
    tuition1 = readr::parse_number(as.character(tuition1))
  )

sfa_sel <- sfa |>
transmute(
unitid,
pct_pell = pgrnt_p # % of undergraduates receiving Pell Grants
)

# ----merge all components--------------------------------------------------

ipeds <- hd |>
  left_join(efia_fte, by = "unitid") |>
  left_join(fin, by = "unitid") |>
  left_join(ic_sel, by = "unitid") |>
  left_join(sfa_sel, by = "unitid") |>
  left_join(adm, by = "unitid") |>
  left_join(om, by = "unitid") |>
  filter(!is.na(unitid), !is.na(fte_total), fte_total > 0) |>
  distinct(unitid, .keep_all = TRUE) |>
  mutate(
    rev_per_fte = tot_rev / fte_total,
    instr_exp_fte = instr_exp / fte_total,
    exp_per_fte = tot_exp / fte_total
  )

# ----summary table (now 'control' exists)---------------------------------
ipeds |>
  summarise(
    total_institutions = n(),
    public_share       = mean(control.y == "Public", na.rm = TRUE),
    nonprofit_share    = mean(control.y == "Private Nonprofit", na.rm = TRUE),
    forprofit_share    = mean(control.y == "Private For-Profit", na.rm = TRUE)
  ) |>
  gt() |>
  fmt_percent(columns = 2:4, decimals = 1)

```

```{r}
## ----inst-type-bar, fig.cap="Distribution of Institutions by Control Type"----
ipeds %>%
  count(control.y) %>%
  ggplot(aes(x = reorder(control.y, n), y = n, fill = control.y)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Distribution of Institutions by Control Type",
    x = "Control Type",
    y = "Number of Institutions"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 13)

## ----tuition-grad-scatter, fig.cap="Tuition vs Graduation Rate by Control Type"----
ipeds %>%
  ggplot(aes(x = tuition1, y = ombach8, color = control.y)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Tuition vs Graduation Rate by Institution Type",
    x = "In-State Tuition (USD)",
    y = "Graduation Rate (%)",
    color = "Control"
  ) +
  theme_minimal(base_size = 13)

## ----geo-dist, fig.cap="Geographic Distribution of Institutions by Control Type"----
ipeds %>%
  ggplot(aes(x = longitud, y = latitude, color = control.y)) +
  geom_point(alpha = 0.5, size = 1.5) +
  coord_map() +
  labs(
    title = "Geographic Distribution of U.S. Institutions",
    x = "Longitude",
    y = "Latitude",
    color = "Control"
  ) +
  theme_minimal(base_size = 13)


```
### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**


### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References

