---
title: "Data-Driven Segmentation of U.S. Colleges and Universities Using K-Means Clustering on IPEDS Data"
author: "Benjamin Minor"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
    link-citations: true       # <-- move here
    link-bibliography: true    # <-- and here
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
nocite: "@*"
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)



## Introduction

Clustering is a central technique in unsupervised machine learning, designed to reveal natural groupings within complex datasets where class labels are unknown. Among its methods, K-Means clustering remains widely applied for its simplicity, scalability, and interpretability [@mohdtalib2023behavioral]. The algorithm partitions data into *k* groups based on similarity, minimizing within-cluster variance while maximizing between-cluster separation. This structure is valuable when the goal is understanding rather than prediction, such as comparing institutions by quantitative indicators of performance, resources, and access.

In higher education research, cluster analysis has long been used to explore structural and performance diversity across institutions [@huberty2005cluster]. This approach enables analyses that move beyond categorical labels and examine measurable dimensions such as outcomes, funding, and enrollment scale. A two-stage framework that combines factor analysis with clustering has also been proposed to derive empirically grounded institutional types [@elam2016classification].

Standardized data and documentation have increased the reliability of these methods [@nces2023groupings]. Best practices for merging and cleaning educational datasets emphasize reproducibility and schema consistency in large-scale analysis [@kelchen2019merging]. Together, these sources point to the value of structured, data-driven methods in institutional research. The Integrated Postsecondary Education Data System (IPEDS), maintained by the National Center for Education Statistics, provides the standardized foundation for this type of large-scale institutional analysis. IPEDS collects consistent and comparable data on U.S. colleges and universities, making it a central source for empirical studies that analyze institutional differences through quantitative modeling.

Hybrid models further advance peer identification by pairing quantitative clustering with expert validation, improving interpretability and practical use [@tair2022peer]. This balance between algorithmic precision and contextual review supports robust benchmarking and decision-making.

The present study applies K-Means clustering to IPEDS data to identify data-driven groupings of postsecondary institutions based on financial, academic, and outcome indicators. The objective is to uncover underlying institutional profiles that describe how U.S. colleges and universities align or differ in measurable ways, offering a transparent and scalable framework for understanding institutional diversity [@huberty2005cluster; @nces2023groupings; @elam2016classification].


## Literature Review

<details>
<summary><strong>Source Summaries (click to expand)</strong></summary>

<div class="markdown" style="margin-top:1rem;">

**Cluster Analysis in Higher Education Research**  
[PDF Link – Springer](https://link.springer.com/content/pdf/10.1007/1-4020-3279-X_8.pdf)  
The chapter reviews how cluster analysis is used to study diversity and structure among higher education institutions. The authors explain how statistical grouping exposes institutional similarities that typical classification systems overlook. They discuss the method’s flexibility for understanding performance and policy patterns across universities and show how it can support empirical research on institutional behavior.  

---

**Institutional Groupings in IPEDS: Considerations for Data Use**  
[Web Link – NCES](https://nces.ed.gov/ipeds/use-the-data/institutional-groupings-in-ipeds)  
This NCES publication outlines the logic and procedures for creating institutional groupings using standardized IPEDS variables. It focuses on the need for consistency, variable alignment, and transparency when comparing data across schools. The document highlights how structured variable relationships can support reproducible and fair institutional comparisons across the national education system.  

---

**Identification of Student Behavioral Patterns in Higher Education Using K-Means Clustering and Support Vector Machine**  
[PDF Link – MDPI Applied Sciences](https://www.mdpi.com/2076-3417/13/5/3267)  
The study applies K-Means clustering to student behavior and performance data, validating results with a Support Vector Machine. The authors used normalized features, centroid checks, and cluster validation to identify three distinct learning behavior groups. They found that clustering helped simplify large datasets into actionable patterns, showing how unsupervised models can classify educational behavior at scale.  

---

**Proposal of a New Method to Classify Higher Education Institutions**  
[PDF Link – Otterbein University Digital Commons](https://digitalcommons.otterbein.edu/cgi/viewcontent.cgi?article=1031&context=stu_dist)  
This thesis introduces a two-step method that combines factor analysis and cluster analysis to build a new classification of universities. The author selected academic, financial, and demographic indicators to find groupings that better represent institutional diversity. The results show that data-driven group formation can outperform fixed classification systems and highlight how numerical traits can define institutional identity.  

---

**A Hybrid Approach to Identifying National Peer Institutions**  
[PDF Link – Texas Association for Institutional Research (TAIR)](https://www.texas-air.org/wp-content/uploads/2022/03/E3-A-Hybrid-Approach-to-Identifying-National-Peer-Institutions.pdf)  
This report explains how statistical clustering and expert review can be combined to identify peer institutions. Quantitative results were first generated from performance and enrollment data, then refined through institutional judgment to improve interpretability. The authors conclude that pairing data models with professional context leads to more balanced and credible peer comparisons.  

---

**Merging Data to Facilitate Analyses**  
[PDF Link – Robert Kelchen](https://robertkelchen.com/wp-content/uploads/2019/11/kelchen_merging_original.pdf)  
Kelchen describes the process of joining and cleaning educational data for large-scale analysis. The guide details how to handle schema differences, missing values, and inconsistent labels across datasets like IPEDS and federal records. It stresses the importance of reproducibility and documentation when merging data sources to maintain accuracy in quantitative research.  


</div>
</details>

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

## Data Description and Visualization

### Dataset Overview

This study uses institutional-level data from the **Integrated Postsecondary Education Data System (IPEDS)** for the **2023 collection year (2023–2024 academic cycle)**, maintained by the **National Center for Education Statistics (NCES)**. IPEDS collects standardized information from all U.S. postsecondary institutions participating in federal financial aid programs. The combined dataset provides national coverage across all control types—public, private nonprofit, and private for-profit—allowing for a comprehensive clustering analysis of institutional characteristics.

The following survey components were used and merged using the common key `UNITID`:

- **HD2023:** Institutional Characteristics (general info, state, sector, control type, FTE enrollment)  
- **F2223_F1A:** Finance – Public institutions (GASB format)  
- **F2223_F2:** Finance – Private nonprofit institutions (FASB format)  
- **F2223_F3:** Finance – Private for-profit institutions (optional inclusion for completeness)  
- **ADM2023:** Admissions (admission rate, SAT/ACT averages, tuition and fees)  
- **OM2023:** Outcome Measures (student completion and success rates)

All data files were downloaded directly from the **NCES IPEDS Data Center**. The merge was performed by `UNITID`, and institutions without valid identifiers or enrollment data were excluded to ensure comparability across survey components.

---

### Variable Selection and Preparation

To capture the main dimensions of institutional performance and scale, the following variables were selected:

| Conceptual Domain | Example Variables (Field Names) |
|--------------------|--------------------------------|
| **Institutional Scale** | `ugds` (undergraduate enrollment), `fte` (full-time equivalent), `student_faculty_ratio` |
| **Financial Resources** | `f1a02` / `f2a02` (total revenues), `f1a13` / `f2a13` (instructional expenses), derived ratios: `revenue_per_fte`, `instruction_exp_per_fte` |
| **Access and Aid** | `admrate` (admission rate), `tuition1` (in-state tuition), `%pell` (Pell Grant recipients) |
| **Outcomes** | `om2_pct_comp` (completion rate), `gradrate` (graduation rate), `retention_rate` |

All numeric variables were standardized using z-score normalization to ensure comparability in K-Means clustering. Outliers were capped at the 1st and 99th percentiles using Winsorization to reduce distortion of centroids.

---

### Data Cleaning and Integration

Public (`F1A`) and private (`F2`) finance datasets were vertically merged using consistent variable naming conventions. For each institution, financial data were joined to demographic and admissions information. Missing values for financial variables were replaced with `NA` rather than imputation, to maintain integrity of original reporting. The merged file included approximately **6,000 active, degree-granting institutions** after filtering.

```{r}
## ----setup, include=FALSE-----------------------------------------------
library(tidyverse)
library(janitor)
library(gt)
library(GGally)
library(scales)
library(here)
library(corrr)

```

```{r}
# Example R pipeline
data_dir <- "data/"

# Institutional characteristics
hd <- read_csv(file.path(data_dir, "HD2023.csv")) |> clean_names()

# Finance: detect public/private files automatically


# Admissions and outcome measures
adm <- read_csv(file.path(data_dir, "adm2023.csv")) |> clean_names()
om  <- read_csv(file.path(data_dir, "om2023.csv"))  |> clean_names()

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References

