---
title: "Data-Driven Segmentation of U.S. Colleges and Universities Using K-Means Clustering on IPEDS Data"
author: "Benjamin Minor"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
    link-citations: true       # <-- move here
    link-bibliography: true    # <-- and here
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
nocite: "@*"
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)



## Introduction

Clustering is a central technique in unsupervised machine learning, designed to reveal natural groupings within complex datasets where class labels are unknown. Among its methods, K-Means clustering remains widely applied for its simplicity, scalability, and interpretability [@mohdtalib2023behavioral]. The algorithm partitions data into *k* groups based on similarity, minimizing within-cluster variance while maximizing between-cluster separation. This structure is valuable when the goal is understanding rather than prediction, such as comparing institutions by quantitative indicators of performance, resources, and access.

In higher education research, cluster analysis has long been used to explore structural and performance diversity across institutions [@huberty2005cluster]. This approach enables analyses that move beyond categorical labels and examine measurable dimensions such as outcomes, funding, and enrollment scale. A two-stage framework that combines factor analysis with clustering has also been proposed to derive empirically grounded institutional types [@elam2016classification].

Standardized data and documentation have increased the reliability of these methods [@nces2023groupings]. Best practices for merging and cleaning educational datasets emphasize reproducibility and schema consistency in large-scale analysis [@kelchen2019merging]. Together, these sources point to the value of structured, data-driven methods in institutional research. The Integrated Postsecondary Education Data System (IPEDS), maintained by the National Center for Education Statistics, provides the standardized foundation for this type of large-scale institutional analysis. IPEDS collects consistent and comparable data on U.S. colleges and universities, making it a central source for empirical studies that analyze institutional differences through quantitative modeling.

Hybrid models further advance peer identification by pairing quantitative clustering with expert validation, improving interpretability and practical use [@tair2022peer]. This balance between algorithmic precision and contextual review supports robust benchmarking and decision-making.

The present study applies K-Means clustering to IPEDS data to identify data-driven groupings of postsecondary institutions based on financial, academic, and outcome indicators. The objective is to uncover underlying institutional profiles that describe how U.S. colleges and universities align or differ in measurable ways, offering a transparent and scalable framework for understanding institutional diversity [@huberty2005cluster; @nces2023groupings; @elam2016classification].


## Literature Review

<details>
<summary><strong>Source Summaries (click to expand)</strong></summary>

<div class="markdown" style="margin-top:1rem;">

**Cluster Analysis in Higher Education Research**  
[PDF Link – Springer](https://link.springer.com/content/pdf/10.1007/1-4020-3279-X_8.pdf)  
The chapter reviews how cluster analysis is used to study diversity and structure among higher education institutions. The authors explain how statistical grouping exposes institutional similarities that typical classification systems overlook. They discuss the method’s flexibility for understanding performance and policy patterns across universities and show how it can support empirical research on institutional behavior.  

---

**Institutional Groupings in IPEDS: Considerations for Data Use**  
[Web Link – NCES](https://nces.ed.gov/ipeds/use-the-data/institutional-groupings-in-ipeds)  
This NCES publication outlines the logic and procedures for creating institutional groupings using standardized IPEDS variables. It focuses on the need for consistency, variable alignment, and transparency when comparing data across schools. The document highlights how structured variable relationships can support reproducible and fair institutional comparisons across the national education system.  

---

**Identification of Student Behavioral Patterns in Higher Education Using K-Means Clustering and Support Vector Machine**  
[PDF Link – MDPI Applied Sciences](https://www.mdpi.com/2076-3417/13/5/3267)  
The study applies K-Means clustering to student behavior and performance data, validating results with a Support Vector Machine. The authors used normalized features, centroid checks, and cluster validation to identify three distinct learning behavior groups. They found that clustering helped simplify large datasets into actionable patterns, showing how unsupervised models can classify educational behavior at scale.  

---

**Proposal of a New Method to Classify Higher Education Institutions**  
[PDF Link – Otterbein University Digital Commons](https://digitalcommons.otterbein.edu/cgi/viewcontent.cgi?article=1031&context=stu_dist)  
This thesis introduces a two-step method that combines factor analysis and cluster analysis to build a new classification of universities. The author selected academic, financial, and demographic indicators to find groupings that better represent institutional diversity. The results show that data-driven group formation can outperform fixed classification systems and highlight how numerical traits can define institutional identity.  

---

**A Hybrid Approach to Identifying National Peer Institutions**  
[PDF Link – Texas Association for Institutional Research (TAIR)](https://www.texas-air.org/wp-content/uploads/2022/03/E3-A-Hybrid-Approach-to-Identifying-National-Peer-Institutions.pdf)  
This report explains how statistical clustering and expert review can be combined to identify peer institutions. Quantitative results were first generated from performance and enrollment data, then refined through institutional judgment to improve interpretability. The authors conclude that pairing data models with professional context leads to more balanced and credible peer comparisons.  

---

**Merging Data to Facilitate Analyses**  
[PDF Link – Robert Kelchen](https://robertkelchen.com/wp-content/uploads/2019/11/kelchen_merging_original.pdf)  
Kelchen describes the process of joining and cleaning educational data for large-scale analysis. The guide details how to handle schema differences, missing values, and inconsistent labels across datasets like IPEDS and federal records. It stresses the importance of reproducibility and documentation when merging data sources to maintain accuracy in quantitative research.  


</div>
</details>

## Methods

This study applies the **K-Means clustering algorithm** to group U.S. higher education institutions into data-driven segments based on scale, resources, access, and outcomes.  
The process follows the standard **Lloyd–Forgy heuristic** for iterative centroid-based optimization, as described in the literature on higher education clustering and institutional classification [@lloyd1982least].

---

### Objective and Notation

Let $X \in \mathbb{R}^{n \times d}$ be the standardized feature matrix, where each row represents an institution and each column a numeric feature (e.g., revenue per FTE, tuition, completion rate).  
The goal of K-Means is to partition $n$ observations into $k$ clusters $C = \{C_1, C_2, \ldots, C_k\}$ and determine centroids $\{\mu_1, \mu_2, \ldots, \mu_k\}$ that **minimize the within-cluster sum of squares (WCSS)**:

$$
\min_{C} \sum_{i=1}^{k} \sum_{x_j \in C_i} \|x_j - \mu_i\|^2
$$

For a fixed cluster assignment, the optimal centroid $\mu_i$ is the arithmetic mean of all observations in $C_i$:

$$
\mu_i = \frac{1}{|C_i|} \sum_{x_j \in C_i} x_j
$$

This optimization problem is **NP-hard** in general; hence, the standard **Lloyd–Forgy iterative heuristic** [@lloyd1982least] is used to approximate convergence through alternating assignment and update steps.


---

### Lloyd’s Algorithm (Assignment–Update Steps)

The algorithm alternates between two main phases until convergence:

1. **Assignment Step:**
   Each observation is assigned to the nearest cluster centroid based on Euclidean distance:
   $$
   C_i^{(t)} = \{x_j : \|x_j - \mu_i^{(t)}\|^2 \leq \|x_j - \mu_r^{(t)}\|^2, \ \forall r \in \{1, \ldots, k\}\}
   $$

2. **Update Step:**
   After assignments, centroids are updated to reflect the mean of their current members:
   $$
   \mu_i^{(t+1)} = \frac{1}{|C_i^{(t)}|} \sum_{x_j \in C_i^{(t)}} x_j
   $$

The process repeats until convergence, typically defined as minimal change in cluster assignments or centroid positions between iterations [@lloyd1982least].

---

### Feature Scaling and Preprocessing

Because IPEDS variables differ in magnitude and measurement units  
(e.g., tuition in USD, completion rate as percentages, and revenue per FTE in dollars),  
all continuous numeric variables were standardized using **z-score normalization** prior to clustering [@lloyd1982least; @huberty2005cluster]:

$$
z = \frac{x - \bar{x}}{s}
$$

where \( \bar{x} \) is the sample mean and \( s \) is the sample standard deviation.  
This ensures that each variable contributes equally to the **Euclidean distance** computation  
used in centroid assignment during K-Means clustering [@lloyd1982least].

Categorical fields, such as institutional control type, were excluded from normalization  
but used later as grouping labels for post-cluster interpretation and comparison.

To mitigate distortion from extreme yet valid institutional profiles (e.g., very large public systems  
or small for-profit colleges), **Winsorization** was applied to cap values beyond the 1st and 99th percentiles.  
This preserves the influence of true outliers while reducing their leverage on centroid placement,  
preventing skew in the cluster boundaries without arbitrarily discarding institutions [@huberty2005cluster].



---

### Implementation Procedure

K-Means was implemented using **R’s base `stats::kmeans()` function** with multiple random initializations (`nstart = 25`) to mitigate sensitivity to initial centroid placement.  
To ensure reproducibility, all random seeds were fixed (`set.seed(2024)`).

Cluster inputs included:

- `REV_PER_FTE` (revenue per full-time equivalent)
- `INSTR_EXP_FTE` (instructional expenses per FTE)
- `TUITION1` (in-state undergraduate tuition)
- `OM8_PCT_COMP` (8-year completion rate)
- `RET_PCF` (retention rate)

The optimal number of clusters $k$ was not pre-specified in this section; it will be **empirically evaluated in the Modeling and Results section** using the **Elbow method** and **Silhouette analysis**, following the recommendations of @huberty2005cluster and @tair2022peer.

---

### Justification of Method

K-Means was selected for its interpretability and computational efficiency on medium-dimensional data.  
Prior studies in higher education analytics have demonstrated that unsupervised clustering effectively reveals institutional typologies that traditional categorical systems overlook [@huberty2005cluster; @nces2023groupings; @mohdtalib2023behavioral; @elam2016classification].  
This approach facilitates **empirical grouping** based on shared performance and resource characteristics, offering actionable insights for benchmarking and peer comparison [@kelchen2019merging].

---

### Cluster Validation and Model Evaluation

To ensure that institutional groupings derived from the K-Means algorithm were both stable and interpretable, multiple internal validation techniques were applied.  
Because K-Means requires the number of clusters \\(k\\) to be specified a priori, identifying an optimal \\(k\\) is a critical methodological step that ensures both interpretability and stability of results [@huberty2005cluster; @tair2022peer].

---

#### Elbow Method (Within-Cluster Sum of Squares)

The **Elbow method** evaluates the total within-cluster sum of squares (WCSS) across a range of \\(k\\) values:

$$
WCSS(k) = \min_C \sum_{i=1}^{k} \sum_{x_j \in C_i} || x_j - \mu_i ||^2
$$

As \\(k\\) increases, WCSS decreases due to smaller cluster radii. The optimal \\(k\\) is typically chosen at the “elbow”: the inflection point beyond which additional clusters yield diminishing reductions in WCSS.  
This heuristic balances compactness (low within-cluster variance) and parsimony (avoiding overfitting through excessive clusters).

---

#### Silhouette Analysis

To complement the Elbow approach, **Silhouette analysis** was used to assess how well each institution fits within its assigned cluster compared to others [@mohdtalib2023behavioral].  
The Silhouette coefficient for each observation \\(i\\) is computed as:

$$
s(i) = \frac{b(i) - a(i)}{\max[a(i),  b(i)]}
$$

where  
- \\(a(i)\\) = average distance between \\(i\\) and all other points in its cluster (cohesion),  
- \\(b(i)\\) = minimum average distance between \\(i\\) and points in the nearest neighboring cluster (separation).  

Silhouette values range from \\(-1\\) to \\(1\\), with higher scores indicating well-separated, coherent clusters.  
The average Silhouette width across all observations was calculated for each candidate \\(k\\), and the \\(k\\) yielding the highest mean silhouette was treated as the most stable partition.

---

#### Model Stability and Reproducibility

To mitigate the randomness inherent in centroid initialization, the clustering routine was executed with multiple random starts (`nstart = 25`) and a fixed seed (`set.seed(2024)`), following best practices for reproducible analysis in R.  
This procedure ensures that the resulting segmentation structure is robust against initialization variance and not an artifact of a single random state [@kelchen2019merging].

---

#### Summary

This validation framework integrates both geometric compactness (Elbow method) and structural separation (Silhouette coefficient) to justify the final number of clusters empirically.  
The actual determination of \\(k\\) and cluster interpretation will be presented in the subsequent **Modeling and Results** section.


## Analysis and Results

### Data Description and Visualization

#### Dataset Overview

This study uses institutional-level data from the **Integrated Postsecondary
Education Data System (IPEDS)** for the 2023 collection year (2023–2024
cycle). IPEDS collects standardized information from all U.S. postsecondary
institutions participating in federal financial aid programs, covering public,
private nonprofit, and private for-profit sectors.

The following survey components were used and merged on the key **`UNITID`**:

- **HD2023** — Institutional characteristics (name, control, sector, etc.)
- **F2223_F1A** — Finance (public, GASB)
- **F2223_F2** — Finance (private nonprofit, FASB)
- **F2223_F3** — Finance (private for-profit)
- **EFIA2023** — 12-month instructional activity (**source of FTE**)
- **ADM2023** — Admissions (admission rate, tuition; optional in analysis)
- **OM2023** — Outcome measures (completion/attainment; optional in analysis)

All files were downloaded from the NCES IPEDS Data Center.

---

#### Variable Selection and Preparation

To represent institutional scale, financial capacity, access, and
outcomes, variables were drawn from multiple components and aligned by
`UNITID`. Below are the primary fields used in this phase:

| Feature Group          | Variable (Field)                                  | Description |
|---|---|---|
| **Institutional Scale** | `FTE_TOTAL` (built from EFIA2023)                 | Total full-time-equivalent enrollment (UG + Grad + DPP if present). |
| **Financial Resources** | `TOT_REV` (F1D01 / F2D162 / F3D12)                 | Total revenues (public), total revenues & investment return (nonprofit/for-profit). |
|                          | `INSTR_EXP` (F1C011 / F2E011 / F3E011)            | Instructional expenses (aligned across control types). |
|                          | `TOT_EXP` (F1N07 / F2B02 / F3B02)                 | Total operating expenses. |
| **Access & Price**       | e.g., `ADM_RATE`, `TUITION1` (ADM2023)            | Admission rate; in-state tuition/fees (if used). |
| **Outcomes**             | e.g., `OM8_PCT_COMP` (OM2023)                     | 8-year outcome/completion measure (if used). |

> Note: Where variables differ by control type, fields were harmonized to a
> shared name (e.g., `TOT_REV`, `INSTR_EXP`, `TOT_EXP`). Ratios per FTE are
> derived after FTE merge.

---

#### Data Cleaning and Integration

Finance data from public (`F1A`), private nonprofit (`F2`), and private
for-profit (`F3`) files were imported and standardized to total revenue,
instructional expense, and total expense. Twelve-month instructional
activity (`EFIA2023`) provided undergraduate, graduate, and total FTE
values; reported FTE were preferred, with estimated values as fallback.
All monetary values are reported in nominal USD for 2023–24 and not 
inflation-adjusted.

Tables were merged using left joins on `UNITID` to preserve coverage. Numeric
missing values were left as `NA` to maintain fidelity with NCES reporting.
After filtering to active, degree-granting institutions with valid FTE, the
analytic file contains ~6,000 institutions across all three control types.

```{r}
# ----libraries-------------------------------------------------------------
library(tidyverse)   # Data wrangling/joins/plots
library(janitor)     # Clean column names
library(gt)          # Summary tables
library(scales)      # Formatting
library(GGally)      # Quick pair plots
library(corrr)       # Correlation summaries
library(here)        # Reproducible relative paths
library(patchwork)   # Grid layouts for multiple plots

# ----directory and base imports-------------------------------------------
data_dir <- here("data")  # adjust if needed

# Core IPEDS components
hd  <- read_csv(file.path(data_dir, "HD2023.csv"))  |> clean_names()
adm <- read_csv(file.path(data_dir, "ADM2023.csv")) |> clean_names()
om  <- read_csv(file.path(data_dir, "OM2023.csv"))  |> clean_names()

# Finance (public / private nonprofit / private for-profit)
f1a <- read_csv(file.path(data_dir, "F2223_F1A.csv")) |> clean_names()
f2  <- read_csv(file.path(data_dir, "F2223_F2.csv"))  |> clean_names()
f3  <- read_csv(file.path(data_dir, "F2223_F3.csv"))  |> clean_names()

# 12-month instructional activity (FTE source)
efia <- read_csv(file.path(data_dir, "EFIA2023.csv")) |> clean_names()

# Student charges (tuition and fees)
ic_ay <- read_csv(file.path(data_dir, "IC2023.csv")) |> clean_names()

# Student financial aid (Pell Grant recipients)
sfa <- read_csv(file.path(data_dir, "SFA2223.csv")) |> clean_names()

# ----build FTE from EFIA2023----------------------------------------------
efia_fte <- efia |>
  mutate(across(c(fteug, efteug, ftegd, eftegd, ftedpp),
                ~ readr::parse_number(as.character(.)))) |>
  transmute(
    unitid,
    fte_ug    = coalesce(fteug, efteug),
    fte_grad  = coalesce(ftegd, eftegd),
    fte_dpp   = ftedpp,
    fte_total = coalesce(fteug, efteug, 0) +
                coalesce(ftegd, eftegd, 0) +
                coalesce(ftedpp, 0)
  )

# ----align finance variables across control types-------------------------
f1a_fin <- f1a |>
  transmute(
    unitid,
    tot_rev   = f1d01,   # Total revenues & other additions (public)
    instr_exp = f1c011,  # Instruction - current year total
    tot_exp   = f1n07,   # Total expenses
    control   = "Public"
  )

f2_fin <- f2 |>
  transmute(
    unitid,
    tot_rev   = f2d162,  # Total revenues & investment return - unrestricted
    instr_exp = f2e011,  # Instruction - total amount
    tot_exp   = f2b02,   # Total expenses
    control   = "Private Nonprofit"
  )

f3_fin <- f3 |>
  transmute(
    unitid,
    tot_rev   = f3d12,   # Total revenues & investment return
    instr_exp = f3e011,  # Instruction - total amount
    tot_exp   = f3b02,   # Total expenses
    control   = "Private For-Profit"
  )

fin <- bind_rows(f1a_fin, f2_fin, f3_fin)

# ----select tuition and Pell variables------------------------------------

ic_sel <- ic_ay |>
  transmute(
    unitid,
    tuition1 = readr::parse_number(as.character(tuition1))
  )

sfa_sel <- sfa |>
transmute(
unitid,
pct_pell = pgrnt_p # % of undergraduates receiving Pell Grants
)

# ----merge all components--------------------------------------------------

ipeds <- hd |>
  left_join(efia_fte, by = "unitid") |>
  left_join(fin, by = "unitid") |>
  left_join(ic_sel, by = "unitid") |>
  left_join(sfa_sel, by = "unitid") |>
  left_join(adm, by = "unitid") |>
  left_join(om, by = "unitid") |>
  filter(!is.na(unitid), !is.na(fte_total), fte_total > 0) |>
  distinct(unitid, .keep_all = TRUE) |>
  mutate(
    rev_per_fte = tot_rev / fte_total,
    instr_exp_fte = instr_exp / fte_total,
    exp_per_fte = tot_exp / fte_total
  )

# ----summary table (now 'control' exists)---------------------------------
ipeds |>
  summarise(
    total_institutions = n(),
    public_share       = mean(control.y == "Public", na.rm = TRUE),
    nonprofit_share    = mean(control.y == "Private Nonprofit", na.rm = TRUE),
    forprofit_share    = mean(control.y == "Private For-Profit", na.rm = TRUE)
  ) |>
  gt() |>
  fmt_percent(columns = 2:4, decimals = 1)

# ----prep-data-------------------------------------------------------------
# Derive 8-year completion rate if missing
if (!"om8_pct_comp" %in% names(ipeds)) {
  need <- c("omawdp8","omnoawd","omenrap","omenryp")
  if (all(need %in% names(ipeds))) {
    ipeds <- ipeds |>
      mutate(
        om_total     = omawdp8 + omnoawd + omenrap + omenryp,
        om8_pct_comp = if_else(om_total > 0, 100 * omawdp8 / om_total, NA_real_)
      )
  }
}

# Parse tuition if text
if ("tuition1" %in% names(ipeds) && !is.numeric(ipeds$tuition1)) {
  ipeds <- ipeds |> mutate(tuition1 = readr::parse_number(as.character(tuition1)))
}

# Prepare data for visuals
vis_df <- ipeds |>
  filter(!is.na(control.y),
         !is.na(tuition1),
         !is.na(om8_pct_comp),
         !is.na(rev_per_fte),
         !is.na(instr_exp_fte)) |>
  mutate(
    tuition_k = tuition1 / 1000,
    rev_fte_k = rev_per_fte / 1000,
    exp_fte_k = instr_exp_fte / 1000
  )

```

#### Figure 1. Distributions of Core Metrics
Compact view of spread/skew for tuition, finance per-FTE, and 8-year completion; motivates standardization before clustering.

```{r}
# ----figure-1--------------------------------------------------------------
#| label: fig-dist
#| fig-cap: "Distributions of Key Variables"
vis_df |>
  select(tuition_k, rev_fte_k, exp_fte_k, om8_pct_comp) |>
  pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "#4682B4", color = "white", alpha = 0.85) +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  labs(x = NULL, y = "Count")

```

#### Figure 2. Tuition vs Completion by Sector
Sector-stratified scatter; dashed global fit shows overall positive slope while sectors vary in level and spread.

```{r}
#----figure-2--------------------------------------------------------------
#| label: fig-tuition-completion
#| fig-cap: "Tuition vs 8-Year Completion by Sector"
ggplot(vis_df, aes(tuition1, om8_pct_comp, color = control.y)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "gray40") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(x = "In-state tuition and fees (USD)", y = "Completion rate (8-year, %)", color = "Sector")
```

#### Figure 3. Revenue vs Instructional Expense per FTE
Bivariate financial profile by sector; public clusters tighter, nonprofits span wider ranges, for-profits trend leaner on expense.

```{r}
#----figure-3--------------------------------------------------------------
#| label: fig-rev-exp-fte
#| fig-cap: "Financial Profile per FTE"
ggplot(vis_df, aes(rev_fte_k, exp_fte_k, color = control.y)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(x = "Revenue per FTE (thousands USD)", y = "Instructional expense per FTE (thousands USD)", color = "Sector")

```

#### Figure 4. Correlation of Key Variables
Heatmap of tuition, finance per-FTE, and completion; confirms finance measures move together and relate to outcomes.

```{r}
#----figure-4--------------------------------------------------------------
#| label: fig-corr
#| fig-cap: "Correlation Heatmap of Tuition, Finances, and Outcomes"
vis_df |>
  select(tuition1, rev_per_fte, instr_exp_fte, om8_pct_comp) |>
    correlate(method = "pearson", use = "pairwise.complete.obs") |>
      stretch(na.rm = TRUE) |>
        ggplot(aes(x, y, fill = r)) +
        geom_tile() +
        geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
        scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
        limits = c(-1, 1), name = "r") +
        labs(x = NULL, y = NULL) +
        theme(axis.text.x = element_text(angle = 40, hjust = 1))

```


### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**


### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References

